Epilepsy example

### 2022-10-12 src/epil/epil.Rmd
Tidy and improve epil notebook

---
title: "Epilepsy example"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
---

@@

@@

@@

@@

## INLA

Both $\tau_\epsilon$ and $\tau_\nu$ have the same $\Gamma(0.001, 0.001)$ prior.
In INLA, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:

```{r}
tau_prior <- list(prec = list(prior = "loggamma",
                              param = c(0.001, 0.001),
                              initial = 1,
                              fixed = FALSE))
```

@@

```{r}
formula <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT +
  f(rand, model = "iid", hyper = tau_prior) +  # Nu random effect
  f(Ind,  model = "iid", hyper = tau_prior)    # Epsilon random effect

epil_inla <- function(strat) {
  inla(formula,
       control.fixed = list(mean = 0, prec = 1/100^2), # Beta prior
       family = "poisson",
       data = Epil,
       control.inla = list(strategy = strat),
       control.predictor = list(compute = TRUE))
}

@@

## TMB

@@

`random` are integrated out with a Laplace approximation:

```{r}
obj <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("epsilon", "nu"),
  DLL = "epil"
)
```

@@

```{r results='hide'}
its <- 1000 #' May converge before this
opt <- nlminb(
  start = obj$par,
  objective = obj$fn,
  gradient = obj$gr,
  control = list(iter.max = its, trace = 0)
)

sd_out <- sdreport(
  obj,
  par.fixed = opt$par,
  getJointPrecision = TRUE
)

#' Check that TMB and Stan have the same objective function for sanity.
#' To obtain the TMB negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
#' For Stan, we create an empty model, then use `rstan::log_prob`.

# tmb_nll <- MakeADFun(data = dat, parameters = param, DLL = "epil") # TMB objective
# stan_nll <- stan("model/epil.stan", data = dat, chains = 0) # Stan objective

#' Now test the NLL of the initialisation parameters `param`:

# c(
#   "TMB" = tmb_nll$fn(unlist(param)),
#   "Stan" = -rstan::log_prob(object = stan_nll, unlist(param))
# )

#' To be more thorough, we can get all the parameters obtained using the MCMC:

# pars_mat <- as.matrix(fit1)
# pars_list <- apply(pars_mat, 1, function(x) relist(flesh = x, skeleton = fit1@inits[[1]]))
# upars_list <- lapply(pars_list, function(x) unconstrain_pars(fit1, x))

#' Just use a few of them

# test_pars <- upars_list[1001:1010]
# tmb_evals <- sapply(test_pars, tmb_nll$fn)
# stan_evals <- -sapply(test_pars, FUN = rstan::log_prob, object = stan_nll)
#
# data.frame(
#   "TMB" = tmb_evals,
#   "Stan" = stan_evals
# )

#' Note: When there are errors in the C++ template code, usually to do with indexing (unlike Stan, in TMB there is no requirement when defining variables to give them dimensions), calling the `MakeADFun` function tends to crash the R session.
#' A workaround (courtesy of Kinh) for debugging without crashing the working R session is to use the following (which creates new R sessions which crash in preference to the working R session).

# library(parallel)
# testrun <- mcparallel({MakeADFun(data = dat,
#                                  parameters = param,
#                                  DLL = "epil")})
# obj <- mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
```

## glmmTMB

[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in TMB without writing the C++ code.
For example, rather than writing `epil.cpp` manually, we could have called the following:

@@

## tmbstan

[`tmbstan`](https://journals.pl,os.org/plosone/article?id=10.1371/journal.pone.0197954) (Cole Monnahan and Kasper Kristensen) is another helpful TMB package which allows you to pass the same C++ template you use in TMB to Stan in order to perform NUTS (if you have standard C++ code then this can also likely be done using [`stanc`](https://statmodeling.stat.columbia.edu/2017/03/31/running-stan-external-c-code/)).

```{r}
fit7 <- tmbstan(obj = obj, chains = 4)
```

## aghq

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer.
One approach is to use `glmmTMB` to get the TMB template which can be passed the `aghq`.

```{r}
# glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)
#
# glmm_ff <- with(glmm_model_info, {
#   TMB::MakeADFun(
#     data = data.tmb,
#     parameters = parameters,
#     random = names(parameters)[grep("theta", names(parameters), invert = TRUE)],
#     DLL = "glmmTMB",
#     silent = TRUE
#   )
# })
#
# glmm_quad <- aghq::marginal_laplace_tmb(glmm_ff, k = 3, startingvalue = glmm_ff$par)
```

@@

## Comparison

@@

@@

Assuming tmbstan to be the ground truth, we can do two dimensional plots to check the fit

```{r}
ggplot(df) +
  geom_point(aes(x = tmbstan, y = TMB, col = type)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5)
```

@@

For `laplace` and `"simplified.laplace"`:

@@

```{r}
plot_marginals <- function(random_effect, index) {
  marginal2 <- fit2$marginals.random[[random_effect]][paste0("index.", index)][[1]]
  marginal3 <- fit3$marginals.random[[random_effect]][paste0("index.", index)][[1]]
  marginal4 <- fit4$marginals.random[[random_effect]][paste0("index.", index)][[1]]

  plot(marginal2, type = "l", col = "red")
  lines(marginal3, col = "green")
  lines(marginal4, col = "blue")
}

>>>>

---
title: "Epilepsy example"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** Rue, Martino and Chopin (2009) use the epilespy example from the OpenBUGs manual as demonstration.

  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan` and `aghq`. We then compare the results of the inference methods.
---

@@

# Background

@@

# Inference methods

@@

Hamiltonian Monte Carlo via the No-U-Turn algorithm, with one Markov chain of `iter = 1000` samples, and `warmup = 100` burn-in:

@@

## `R-INLA`

Both $\tau_\epsilon$ and $\tau_\nu$ have the same $\Gamma(0.001, 0.001)$ prior.
In `R-INLA`, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:

```{r}
tau_prior <- list(prec = list(
  prior = "loggamma",
  param = c(0.001, 0.001),
  initial = 1,
  fixed = FALSE)
)
```

@@

```{r}
formula <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT +
  f(rand, model = "iid", hyper = tau_prior) +  # Nu random effect
  f(Ind,  model = "iid", hyper = tau_prior)    # Epsilon random effect
```

We can use various settings for the strategy, which controls how the latent field posterior marginals are approximated:

```{r}
epil_inla <- function(strat) {
  inla(formula,
       control.fixed = list(mean = 0, prec = 1/100^2), # Beta prior
       family = "poisson",
       data = Epil,
       control.inla = list(strategy = strat),
       control.predictor = list(compute = TRUE))
}

@@

## `TMB`

@@

We integrate `random` out with a Laplace approximation:

```{r}
obj <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("epsilon", "nu"),
  DLL = "epil",
  silent = TRUE
)
```

@@

```{r}
its <- 1000 #' May converge before this

opt <- nlminb(
  start = obj$par,
  objective = obj$fn,
  gradient = obj$gr,
  control = list(iter.max = its, trace = 0)
)

sd_out <- sdreport(
  obj,
  par.fixed = opt$par,
  getJointPrecision = TRUE
)
```

For completeness, we check here that the objective functions as implemented in Stan and `TMB` are identical.
To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
For Stan, we create an empty model, then use `rstan::log_prob`:

```{r}
tmb_nll <- MakeADFun(data = dat, parameters = param, DLL = "epil")
stan_nll <- stan("epil.stan", data = dat, chains = 0)
```

Now test the NLL of the initialisation parameters `param`:

```{r}
c(
  "TMB" = tmb_nll$fn(unlist(param)),
  "Stan" = -rstan::log_prob(object = stan_nll, unlist(param))
)
```

To be more thorough, we can get all the parameters obtained using the MCMC:

```{r}
pars_mat <- as.matrix(fit1)
pars_list <- apply(pars_mat, 1, function(x) relist(flesh = x, skeleton = fit1@inits[[1]]))
upars_list <- lapply(pars_list, function(x) unconstrain_pars(fit1, x))
```

Using just a few of them, we can see that there is a difference, but that difference is a constant so doesn't matter:

```{r}
test_pars <- upars_list[500:510]
tmb_evals <- sapply(test_pars, tmb_nll$fn)
stan_evals <- -sapply(test_pars, FUN = rstan::log_prob, object = stan_nll)

data.frame(
  "TMB" = tmb_evals,
  "Stan" = stan_evals
) %>%
  mutate(Difference = TMB - Stan)
```

Note: When there are errors in the C++ template code, usually to do with indexing (unlike Stan, in `TMB` there is no requirement when defining variables to give them dimensions), calling the `MakeADFun` function tends to crash the R session.
A workaround (courtesy of Kinh) for debugging without crashing the working R session is to use the following (which creates new R sessions which crash in preference to the working R session):

```{r eval=FALSE}
library(parallel)

testrun <- mcparallel({MakeADFun(data = dat,
                                 parameters = param,
                                 DLL = "epil")})

obj <- mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
```

## `glmmTMB`

[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in `TMB` without writing the C++ code.
For example, rather than writing `epil.cpp` manually, we could have called the following:

@@

This likely isn't the same model, in terms of priors and so on, but it may be close.

## `tmbstan`

[`tmbstan`](https://journals.pl,os.org/plosone/article?id=10.1371/journal.pone.0197954) (Cole Monnahan and Kasper Kristensen) is another helpful `TMB` package which allows you to pass the same C++ template you use in `TMB` to Stan in order to perform NUTS (if you have standard C++ code then this can also likely be done using [`stanc`](https://statmodeling.stat.columbia.edu/2017/03/31/running-stan-external-c-code/)).

```{r}
fit7 <- tmbstan(obj = obj, chains = 1)
```

## `aghq`

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
One (strange!) approach is to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.

```{r}
glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)

glmm_ff <- with(glmm_model_info, {
  TMB::MakeADFun(
    data = data.tmb,
    parameters = parameters,
    random = names(parameters)[grep("theta", names(parameters), invert = TRUE)],
    DLL = "glmmTMB",
    silent = TRUE
  )
})
```

The parameter `k = 3` determines the number of quadrature points per dimension:

```{r}
glmm_quad <- aghq::marginal_laplace_tmb(glmm_ff, k = 3, startingvalue = glmm_ff$par)
```

@@

# Comparison

## Table

@@

kableExtra::kable(df)

@@

## Scatter plot against `tmbstan`

Assuming `tmbstan` to be the ground truth, we can do two dimensional plots to check the fit

```{r}
ggplot(df) +
  geom_point(aes(x = tmbstan, y = TMB, col = type)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
  theme_minimal()
```

## Symmetric KL divergence

@@

For `laplace` and `"simplified.laplace"` we can look for the random effect with the highest SKLD:

@@

```{r}
plot_marginals <- function(random_effect, index) {
  marginal2 <- fit2$marginals.random[[random_effect]][paste0("index.", index)][[1]] %>%
    as.data.frame() %>%
    mutate(method = "G")

  marginal3 <- fit3$marginals.random[[random_effect]][paste0("index.", index)][[1]] %>%
    as.data.frame() %>%
    mutate(method = "SL")

  marginal4 <- fit4$marginals.random[[random_effect]][paste0("index.", index)][[1]] %>%
    as.data.frame() %>%
    mutate(method = "L")

  marginals <- bind_rows(marginal2, marginal3, marginal4)

  ggplot(marginals, aes(x = x, y = y, group = method, col = method)) +
    geom_line() +
    theme_minimal()
}
